The Design and Implementation of the Synchronous Language \CEU

### 2016-04-06 paper.tex
...

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with simpler semantics and 
more fine-grained control over program execution.
%
\CEU adopts an event-triggered notion of time that enables compile-time 
temporal analysis to detect conflicting statements, resulting in deterministic 
and concurrency-safe programs.
%
Using Esterel as a base, we discuss the particularities of our design such as 
stack-based internal events, the temporal analysis, safe integration with \C, 
and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
\end{abstract}

@@

Despite its success and influence, Esterel has an overly complex semantics that 
requires careful static analysis such as to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems.
These problems have an extensive coverage in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible and non-compliant implementations.
%
Above all, it also affects the programmer's understanding about the code, 
which, ultimately, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of loose and 
non-deterministic execution for intra-reaction statements.
Non determinism prevents threads to interact with stateful system calls safely 
and makes shared-memory concurrency not as straightforward as reading and 
writing to shared variables.
%
%For instance, although Esterel supports orthogonal abortion of 
%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
%release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
% TODO: simple simple
Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution, and is amenable to a temporal analysis that
improves safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C which enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of our simple semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

In the subsections that follow, we discuss the main differences between \CEU 
and Esterel:
%
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
We finish with a summary of our design
    (\ref{sec.ceu.discussion}).
%
We provide the formal specification of the semantics of \CEU in a separate 
work~\cite{ceu.phd}.

@@

Esterel defines time as a discrete sequence of logical unit instants or 
``ticks''.
At each tick, the program reacts to an arbitrary number of simultaneous input 
events, depending on external stimuli the environment provides.
In contrast, \CEU defines time as a discrete sequence of reactions to
unique input events.
At each input event, which constitutes a logical unit of time, the program 
reacts exclusively to it.
The event-triggered execution of a program in \CEU is as 
follows~\cite{ceu.sensys13}:
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail (but 
      parallel constructs may create new trails).
\item Active trails execute until they await or terminate.
      This step is named a \emph{reaction chain}, and always runs in bounded 
      time.
\item The program goes idle and the environment takes control.
\item On the occurrence of a new external input event, the environment awakes 
      \emph{all} trails awaiting that event.
      It then goes to step 2.
\end{enumerate}
%
A program must react to an occurring event completely before handling the next.
%
Based on the synchronous hypothesis, a program takes a negligible time on step 
2 and is always idle on step 3.
In practice, if a new external input event occurs while a reaction chain is 
running, it is enqueued to occur in the next reaction.

@@

Figure~\ref{fig.ticks} compares the discrete notions of time in two variations 
of Esterel and in \CEU.
The box \code{Real World} shows event occurrences over a continuous timeline 
divided in units of 10 milliseconds.
The other boxes define distinct discrete logical units of time in which the 
same occurring events fit differently.
%
\begin{itemize}
\item \code{[Box-1]}:
``Esterel with fixed-length ticks'',
within which reactions must fit~\cite{esterel.fixed}.
%
We assume a \code{R(boot)} reaction at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes in the next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1}, causing a 
\emph{timing violation}~\cite{esterel.fixed} that invades \code{tick-2}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur during the idle \code{tick-3}.
However, only one occurrence of \code{E} is considered in \code{R(D+E)}.
%
\item \code{[Box-2]}:
``Esterel with variable-length ticks'', which are adjusted to the CPU time the 
corresponding reaction takes to complete~\cite{esterel.variable}.
This approach avoids the time violation for \code{R(A)} and also results in
smaller idle periods.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
react alone as \code{R(D)} on \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered on \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{[Box-3]}:
``\CEU with unique and queue-based input events''.
We also assume a \code{R(boot)} reaction before any input.
%
Because the occurrence of event \code{A} is unique during \code{tick-0}, the 
behavior in \CEU is similar to \code{Box-2} for the first two units of time 
(\code{boot-0} and \code{A-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} as 
simultaneous, and handles each in subsequent reactions \code{R(B)} and 
\code{R(C)}.
%
We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
\CEU to be roughly the same.
This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU recognizes and reacts to the two instances of \code{E} 
independently, which are handled in sequence.
\end{itemize}

We decided for the unique and queue-based semantics in \CEU for the reasons 
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never absolutely simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but defined explicitly for each use case.
In the case of Esterel, simultaneity depends on the imprecise length of 
discrete ticks.
For instance, in Figure~\ref{fig.ticks} (\code{box-1} and \code{box-2}), the 
events \code{B} and \code{C} are considered simultaneous, even though \code{A} 
and \code{B} ``physically'' happen much closer to one another.
In Section~\ref{sec.ceu.timers}, after introducing internal events and 
first-class timers, we show an example to detect simultaneous button clicks in 
\CEU.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to multiple events never overlap because they are atomic.
Automated mutual exclusion simplifies the reasoning about the program and is a 
prerequisite for the temporal analysis to be discussed in 
Section~\ref{sec.ceu.analysis}.
\end{itemize}

@@

Esterel makes no semantic distinctions between internal and external signals.
%In particular, programs can emit external and internal signals simultaneously, 
%with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
%In \CEU, subsequent external events coming from the environment define a 
%timeline, which programs cannot interfere by emitting new external events.
%
%To cope with deterministic intra-reaction communication, \CEU supports 
%internal events which programs can \code{emit} and \code{await}.
In \CEU, in contrast with queue-based external events, internal events follow a 
stack-based execution policy similar to subroutine calls in typical programming 
languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel \ax and \CEU \bx.
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
(ln. 6,9) in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes, broadcasts \code{b}, and pauses (ln. 4--5).
\item 2nd trail awakes, calls \code{\_g()}, and terminates (ln. 8--9).
      \emph{(No other trails awake to \code{b}.)}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates (ln. 5--6).
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

Internal events provide a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed as 
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (ln. 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence the calling trail resumes and passes the assertion 
test (ln. 11).

\CEU also supports nested \code{emit} invocations for internal events.
For instance, the code that executes after the \code{await inc} (ln. 4) could 
\code{emit} another event, creating a new level in the stack.
The runtime stack represents fine-grained micro reactions, one on top of the 
other, all inside the same reaction to an external event.
%Internal events provide a fine-grained intra-reaction control mechanism.

@@

\begin{figure}[b]
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e;     // w/o delayed awaits, the emit awakes 1st trail
        await f;    // and restarts the loop instantaneously
    end
end
\end{lstlisting}
\caption{ Delayed awaits prevents re-execution of statements by design.
\label{lst.await}
}
\end{figure}

The example in Figure~\ref{lst.await} illustrates delayed awaits, which 
prevents infinite execution by design.
Both sides of the \code{par/or} have an \code{await} statement to avoid 
instantaneous termination (ln. 4,7) of the enclosing \code{loop} (ln 2--9).
However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
same reaction that reaches them, the \code{par/or} would terminate and restart 
the \code{loop} instantaneously, resulting in infinite execution.
%
\begin{comment}
\begin{figure}
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits some period
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute immediately and then periodically
        await e;  // await after
    end
end
\end{lstlisting}
%\end{minipage}
%
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[numbers=left,xleftmargin=3em]
%event void e;
%var bool should_execute = false;
%par do
%    loop do
%        <...>         // code that awaits
%        if <...> then // some conditon
%            should_execute = true;
%            emit e;
%        end
%    end
%with
%    loop do
%        if should_execute then
%            <...>     // code to execute
%        end
%        await e;
%    end
%end
%\end{lstlisting}
%\end{minipage}
%
\caption{ An example that circumvents the \emph{delayed await} by post-fixing 
the \code{await} inside the \code{loop}.
% (in the left), and by copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}
\end{comment}
%
In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
\begin{comment}
For instance, sometimes we need to execute a block of code immediately, and 
then periodically from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
\end{comment}
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for complex analysis to deal with schizophrenic statements.

@@

Esterel is only deterministic with respect to external behavior: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
|| call f2()>>}, the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
A number of Esterel-based synchronous languages enforce intra-reaction 
determinism with an arbitrary execution order for statements in multiple lines 
of execution (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%, e.g., based on lexical order or manual priority assignment.
%
\CEU also takes the deterministic approach and, when multiple trails are active 
during the same reaction, they are scheduled in lexical order, i.e., in the 
order they appear in the program source code.

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax Accesses to \code{x} are safe}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\centering\small{\bx Accesses to \code{y} are unsafe}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
Example \ax is safe because the trails access \code{x} atomically in different 
reactions;
Example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure}

Even so, we consider that enforcing an arbitrary execution order can be 
misleading in some cases.
%
For instance, consider the two examples in Figure~\ref{lst.shared}, both of 
which define a shared variable (ln. 2), and assign to it in parallel trails 
(ln.  5, 8).
%
In the example \ax, the two assignments to \code{x} only execute in reactions 
to different events \code{A} and \code{B} which, by definition, cannot occur 
simultaneously (Section~\ref{sec.ceu.exts}).
Hence, for the sequence \code{A->B}, \code{x} becomes \code{4=(1+1)*2}, while 
for \code{B->A}, \code{x} becomes \code{3=(1*2)+1}.
%
In the example \bx, the two assignments to \code{y} are simultaneous because 
they execute in reaction to the same event \code{A}.
Since \CEU employs lexical order for intra-reaction statements, the execution 
is still deterministic, and \code{y} always becomes \code{4=(1+1)*2}.
%
However, an (apparently innocuous) change in the order of trails modifies the 
semantics of the program, which we consider unsafe.

>>>>

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with simpler semantics and 
more fine-grained control over program execution.
%
\CEU adopts an event-triggered notion of time that enables compile-time 
temporal analysis to detect conflicting statements, resulting in deterministic 
and concurrency-safe programs.
%
Using Esterel as a base, we discuss the particularities of our design, such as 
stack-based internal events, the temporal analysis, safe integration with \C, 
and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
\end{abstract}

@@

Despite its success and influence, Esterel has an overly complex semantics that 
requires careful static analysis such as to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems%
%These problems have an extensive coverage in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible and non-compliant implementations.
%
Above all, it also affects the programmer's understanding about the code, 
which, ultimately, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of loose and 
non-deterministic execution for intra-reaction statements.
Non determinism prevents threads to interact with stateful system calls safely 
and makes shared-memory concurrency not as straightforward as reading and 
writing to shared variables.
%
%For instance, although Esterel supports orthogonal abortion of 
%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
%release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
% TODO: simple simple
Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution, and is amenable to a temporal analysis that
improves safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C, which enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of our simple semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

In the subsections that follow, we discuss the main differences between \CEU 
and Esterel:
%
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
We finish with a summary of our design
    (\ref{sec.ceu.summary}).
%
We present the formal specification of the semantics of \CEU in a separate 
work~\cite{ceu.phd}.

@@

Esterel defines time as a discrete sequence of logical unit instants or 
``ticks''.
At each tick, the program reacts to an arbitrary number of simultaneous input 
events from the environment.
In contrast, \CEU defines time as a discrete sequence of reactions to
unique input events.
At each input event, which constitutes a logical unit of time, the program 
reacts exclusively to it.
The event-triggered execution of a program in \CEU is as 
follows~\cite{ceu.sensys13}:
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail (but 
      parallel constructs may create new trails).
\item Active trails execute, one after the other, until they await or 
      terminate.
      This step is named a \emph{reaction chain}, and always runs in bounded 
      time.
\item The program goes idle and the environment takes control.
\item On the occurrence of a new external input event, the environment awakes 
      \emph{all} trails awaiting that event.
      It then goes to step 2.
\end{enumerate}
%
A program must react to an occurring event completely before handling the next.
%
Based on the synchronous hypothesis, a program takes a negligible time on step 
2 and is always idle on step 3.
In practice, if a new external input event occurs while a reaction chain is 
running, it is enqueued to occur in the next reaction.

@@

Figure~\ref{fig.ticks} compares the discrete notions of time in two variations 
of Esterel and in \CEU.
The box \code{Real World} shows event occurrences over a continuous timeline.
The other boxes show discrete logical units of time in which the same occurring 
events fit in different ways.
%
\begin{itemize}
\item \code{[Box-1]}:
\emph{Esterel with fixed-length ticks},
in which reactions must fit~\cite{esterel.fixed}.
%
We assume a reaction \code{R(boot)} at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes in the next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1}, causing a 
\emph{timing violation}~\cite{esterel.fixed} that invades \code{tick-2}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur during the idle \code{tick-3}.
However, only one occurrence of \code{E} is considered in \code{R(D+E)}.
%
\item \code{[Box-2]}:
\emph{Esterel with variable-length ticks}, which are adjusted to the CPU time 
the corresponding reaction takes to complete~\cite{esterel.variable}.
This approach avoids the time violation for \code{R(A)} and also results in
smaller idle periods.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
react alone as \code{R(D)} on \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered on \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{[Box-3]}:
\emph{\CEU with unique and queue-based input events}.
We also assume a reaction \code{R(boot)} before any input.
%
Because the occurrence of event \code{A} is unique during \code{tick-0}, the 
behavior in \CEU is similar to \code{Box-2} for the first two reactions 
(\code{boot-0} and \code{A-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} as 
simultaneous, and handles each in subsequent reactions \code{R(B)} and 
\code{R(C)}.
%
We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
\CEU to be roughly the same.
This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU recognizes and reacts to the two instances of \code{E} 
independently, which are handled in sequence.
\end{itemize}

We decided for the unique and queue-based semantics in \CEU for the reasons 
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never absolutely simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but defined explicitly for each use case.
In the case of Esterel, simultaneity depends on the imprecise length of 
discrete ticks.
For instance, in Figure~\ref{fig.ticks} (\code{box-1} and \code{box-2}), 
Esterel considers the events \code{B} and \code{C} to be simultaneous, even 
though \code{A} and \code{B} ``physically'' happen much closer to one another.
In Section~\ref{sec.ceu.timers}, after introducing internal events and 
first-class timers, we show an example to detect simultaneous button clicks in 
\CEU.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to multiple events in \CEU never overlap because they are atomic.
Automated mutual exclusion simplifies the reasoning about concurrency and is a 
prerequisite for the temporal analysis to be discussed in 
Section~\ref{sec.ceu.analysis}.
\end{itemize}

@@

Esterel makes no semantic distinctions between internal and external signals.
%In particular, programs can emit external and internal signals simultaneously, 
%with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
%In \CEU, subsequent external events coming from the environment define a 
%timeline, which programs cannot interfere by emitting new external events.
%
%To cope with deterministic intra-reaction communication, \CEU supports 
%internal events which programs can \code{emit} and \code{await}.
In \CEU, in contrast with queue-based external events, internal events follow a 
stack-based execution policy similar to subroutine calls in typical programming 
languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel \ax and \CEU \bx.
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order (ln. 6,9).
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes, broadcasts \code{b}, and pauses (ln. 4--5).
\item 2nd trail awakes, calls \code{\_g()}, and terminates (ln. 8--9).
      \emph{(No other trails awake to \code{b}.)}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates (ln. 5--6).
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

Internal events provide a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed as 
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (ln. 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence, the calling trail resumes and passes the assertion 
test (ln. 11).

\CEU also supports nested \code{emit} invocations for internal events.
For instance, the code in Figure~\ref{lst.sub} that executes after the 
\code{await inc} (ln.  4) could \code{emit} another event, creating a new level 
in the stack.
The runtime stack constitutes fine-grained micro reactions, one on top of the 
other, all inside the same reaction to an external event.
%Internal events provide a fine-grained intra-reaction control mechanism.

@@

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e;     // w/o delayed awaits, the emit awakes 1st trail
        await f;    // and restarts the loop instantaneously
    end
end
\end{lstlisting}
\caption{ Delayed awaits prevents re-execution of statements by design.
\label{lst.await}
}
\end{figure}

The example in Figure~\ref{lst.await} illustrates delayed awaits, which 
prevents infinite execution by design.
Both sides of the \code{par/or} have an \code{await} statement (ln. 4,7), which 
makes the enclosing \code{loop} non instantaneous (ln 2--9).
However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
same reaction that reaches them, the \code{par/or} would terminate and restart 
the \code{loop} instantaneously, resulting in infinite execution.
%
\begin{comment}
\begin{figure}
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits some period
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute immediately and then periodically
        await e;  // await after
    end
end
\end{lstlisting}
%\end{minipage}
%
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[numbers=left,xleftmargin=3em]
%event void e;
%var bool should_execute = false;
%par do
%    loop do
%        <...>         // code that awaits
%        if <...> then // some conditon
%            should_execute = true;
%            emit e;
%        end
%    end
%with
%    loop do
%        if should_execute then
%            <...>     // code to execute
%        end
%        await e;
%    end
%end
%\end{lstlisting}
%\end{minipage}
%
\caption{ An example that circumvents the \emph{delayed await} by post-fixing 
the \code{await} inside the \code{loop}.
% (in the left), and by copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}
\end{comment}
%
In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
\begin{comment}
For instance, sometimes we need to execute a block of code immediately, and 
then periodically from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
\end{comment}
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for complex analysis to deal with schizophrenic statements.

@@

Esterel is only deterministic with respect to external behavior: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
|| call f2()>>}, the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
A number of Esterel-based synchronous languages have intra-reaction determinism 
by enforcing an arbitrary execution order for statements in multiple lines of 
execution (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%, e.g., based on lexical order or manual priority assignment.
%
\CEU also takes the deterministic approach and, when multiple trails are active 
during the same reaction, they are scheduled in lexical order, i.e., in the 
order they appear in the program source code.

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax Accesses to \code{x} are safe}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\centering\small{\bx Accesses to \code{y} are unsafe}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure}

Even so, we consider that enforcing an arbitrary execution order can be 
misleading in some cases.
%
For instance, consider the two examples in Figure~\ref{lst.shared}, both 
defining a shared variable (ln. 2), and assigning to it in parallel trails (ln.  
5, 8).
%
In the example \ax, the two assignments to \code{x} execute in reactions to 
different events \code{A} and \code{B} only, which by definition, cannot occur 
simultaneously (Section~\ref{sec.ceu.exts}).
Hence, for the sequence \code{A->B}, \code{x} becomes \code{4} 
(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes \code{3} 
(\code{(1*2)+1}).
%
In the example \bx, the two assignments to \code{y} are simultaneous because 
they execute in reaction to the same event \code{A}.
Since \CEU employs lexical order for intra-reaction statements, the execution 
is still deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).
%
However, an (apparently innocuous) change in the order of trails modifies the 
semantics of the program, which we consider unsafe.
